\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-09 k-planar.tex
started appendix

\end{document}s

>>>>

\appendix
\section{New Stuff}

\note{DW}{Work in progress}

For a plane multigraph $G$, let $\widehat{G}$ be the embedded graph obtained from $G$ as follows: for each face $f$ of $G$ and for each pair of non-consecutive vertices $v$ and $w$ on the boundary of $f$, add the edge $vw$ to $\widehat{G}$ drawn across $f$.

\begin{thm}
\label{CliqueFaceProductStructure}
Let $G$ be a 2-connected plane multigraph, in which every facial cycle has length in $\{3,\dots,d\}$ for some fixed integer $d\geq 3$. Then $\widehat{G}$ is contained in $H \boxtimes P \boxtimes K_{ d  \floor{d/2} }$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

Note that \cref{CliqueFaceProductStructure} with $d=3$ implies \cref{PlanarProduct}(b)  by \citet{DJMMUW20} (since every planar graph is a subgraph of a triangulation). Moreover, the proof of \cref{CliqueFaceProductStructure} is a direct generalisation of the proof of 
\cref{PlanarProduct}(b).

For a cycle $C$, we write $C = [ P_1 , \dots , P_k ]$ if $P_1 , \dots, P_k$ are pairwise disjoint non-empty paths in $C$, and the endpoints of each path $P_i$ can be labelled $x_i$ and $y_i$ so that $y_ix_{i+1} \in E(C)$ for $i\in\{1,\dots,k\}$, where $x_{k+1}$ means
$x_1$. This implies that $\{P_1,\dots,P_k\}$ is a vertex-partition of $C$.

If $Q$ is a path in a graph, and $P_1, \dots,P_k$ are edge-disjoint non-empty subpaths of $Q$ ordered by $Q$ and partitioning $E(Q)$, then we write $Q=P_1,\dots,P_k$.

We employ the following well-known variation of Sperner's Lemma (see~\citep{Proofs4}).

\begin{lem}[Sperner's Lemma]
\label{Sperner}
Let $G$ be a near-triangulation whose vertices are coloured $1,2,3$, with the outer-face $F=[P_1,P_2,P_3]$ where each vertex in $P_i$ is coloured $i$. Then $G$ contains an internal face whose vertices are coloured $1,2,3$. \note{DW}{Does this allow parallel edges: check the proof in \citep{Proofs4}.}
\end{lem}

%If $C$ is a cycle in a graph, and $P_1, \dots,P_k$ are edge-disjoint non-empty subpaths of $C$ ordered around $C$ and partitioning $E(C)$, then we write $C=(P_1,\dots,P_k)$.

Say $G$ is a 2-connected plane multigraph. Let $T$ be a spanning tree of $G$ rooted at a vertex $r$ on the outerface of $G$. A \defin{tripod} (with respect to $T$) is any subgraph of $G$ consisting of three vertical paths $P,Q,R$ in $T$ plus a facial cycle $F$ of $G$,
where the lower endpoint of each of $P,Q,R$ is in $F$. A \defin{bipod} (with respect to $T$) is any (connected) subpath of a tripod (with respect to $T$).
 %is a path $PXQ$ where $P$ and $Q$ are vertical paths in $T$ and $X$ is a facial path of $G$, where the lower endpoints of $P$ and $Q$ are the endpoints of $X$. 
\note{DW}{Need to think about degenerate cases. We want every bipod to be a tripod}
%, and every connected subgraph of a bipod to be a bipod.}.
 These definitions  generalise definitions in \citep{DJMMUW20} where $G$ is assumed to be a plane triangulation, in which case $F$ is a 3-cycle and $X$ is an edge. \note{DW}{We actually managed to omit the definition of `bipod' in \citep{DJMMUW20}!}

%In the original proof for planar triangulation, B_i is an edge.

\begin{lem}
Fix an integer $d \geq 3$. Say $G_0$ is a 2-connected plane multigraph, in which each facial cycle has length in $\{3,\dots,d\}$. \note{DW}{I expect $d$ will be irrlevant to this lemma.} Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $k\in\{1,2,3\}$ and let $C = [A_1,\dots,A_k]$ be a cycle in $G_0$ with $r$ is in the exterior of $C$, where 
$A_1, \dots, A_k$ are bipods in $G_0$. Let $G$ be the plane graph consisting of all the edges and vertices of $G_0$ in $C$ and the interior of $C$.
Then $G$ has a partition $\PP$ into tripods with resect to $T$ such that $P_1,\dots,P_k\in\PP$ and the graph $H : = G / \PP$ is planar and has a tree-decomposition in which every bag has size at most 4 and some bag contains all the vertices of $H$ corresponding to $P_1,\dots,P_k$. \note{DW}{Do we need to add that ``each face of $G_0$ is the centre of some tripod in $\PP$'' or something like that? Then each facial clique will be in some part.}
\end{lem}

\begin{proof}
	\note{DW}{Is $C$ allowed to be a 2-cycle?}
The proof is by induction on $|V(G)\setminus V(C)|$. If $V(G)\setminus V(C)=\emptyset$, then the desired partition is $\PP=\{P_1,\ldots,P_k\}$.  The tree-decomposition of $H$ consists of a single bag that contains the $k\le 3$ vertices corresponding to $P_1,\ldots,P_k$.

Now assume that $V(G)\setminus V(C)\neq\emptyset$. Partition the vertices of $C$ into paths $R_1$, $R_2$, and $R_3$ as follows:
\begin{compactenum}
	\item If $k=1$ then, since $C$ is a cycle, $P_1$ has at least three vertices, so $P_1=[v, P_1', w]$ for two distinct vertices $v$ and $w$.   \note{DW}{need to be careful with parallel edges} We set $R_1:=v$, $R_2:=P_1'$ and $R_3:=w$.

	\item If $k=2$ then we may assume without loss of generality that $P_1$ has at least two vertices so $P_1=[v,P_1']$ \note{DW}{need to be careful with parallel edges}.  We set $R_1:=v$, $R_2:=P_1'$ and $R_3:=P_2$.

	\item If $k=3$ then let $R_i:=P_i$ for each $i\in\{1,2,3\}$.
\end{compactenum}

Let $G'$ be an arbitrary triangulation of $G_0$. For $i\in\{1,2,3\}$, colour each vertex in $R_i$ by $i$. Now, for each remaining vertex $v$ in $G$, consider the path $P_v$ from $v$ to the root of $T$. Since $r$ is on the outer-face of $G_0$, $P_v$ contains at least one vertex of $C$. If the first vertex of $P_v$ that belongs to $F$ is in $R_i$ then assign the colour $i$ to $v$. We obtain a 3-colouring of the vertices of $G'$ that satisfies the conditions of Sperner's Lemma. By Sperner's Lemma there exists a triangular face $\tau=v_1v_2v_3$ of $G'$ whose vertices are coloured $1,2,3$ respectively. Let $F$ be the face of $G_0$ containing $\tau$.

For each $i\in\{1,2,3\}$, let $Q_i$ be the path in $T$ from $v_i$ to the first ancestor $v_i'$ of $v_i$ in $T$ that is contained in $C$. Observe that $Q_1$, $Q_2$, and $Q_3$ are disjoint since $Q_i$ consists only of vertices coloured $i$. Note that $Q_i$ may consist of the single vertex $v_i=v_i'$.  Let $Q_i'$ be $Q_i$ minus its final vertex $v_i'$.  Imagine for a moment that the cycle $C$ is oriented clockwise, which defines an orientation of $R_1$, $R_2$ and $R_3$. Let $R_i^-$ be the subpath of $R_i$ that contains $v'_i$ and all vertices that precede it, and let $R_i^+$ be the subpath of $R_i$ that contains $v'_i$ and all vertices that succeed it.

Say $F=v_1 S_1 v_2 S_2 v_3 S_3$, where $S_i$ is a facial path in $F$ clockwise from $v_i$ to $v_{i+1}$.

, where we use the convention that $Q_4=Q_1$ and $R_4=R_1$. Note that $C_i$ may be \emph{degenerate} in the sense that $[Q_i',R_i^+,R_{i+1}^-,Q_{i+1}']$ may consist only of a single edge $v_iv_{i+1}$.

Let $Y$ be the tripod consisting of $Q_1'\cup Q_2' \cup Q_3'$ plus the vertices and edges of $F$ between non-empty $Q_1',Q_2',Q_3'$.

\note{DW}{Much of the following paragraph can be deleted} Consider any non-degenerate $C_i=[Q_i',R_i^+,R_{i+1}^-,Q_{i+1}']$. Note that these four paths  are pairwise disjoint, and thus $C_i$ is a cycle. If  $Q_i'$ and $Q_{i+1}'$ are non-empty, then each is a vertical path in $T$.  Furthermore, each of $R_i^-$ and $R_{i+1}^+$ consists of at most two vertical paths in $T$.  Thus, $C_i$ is the concatenation of at most six vertical paths in $T$ \note{DW}{re-do}.  Let $G_i$ be the subgraph of $G_0$ consisting of all the edges and vertices of $G_0$ contained in $C_i$ and the interior of $C_i$.  Observe that $G_i$ contains $v_i$ and $v_{i+1}$ but not the third vertex of $\tau$.

Consider the graph consisting of the edges and vertices of $F$, the edges and vertices of $C$, and the edges and vertices of $Q_1,Q_2,Q_3$. This graph has up to three \note{DW}{can be more} internal faces $C_1,C_2,C_3$ where each $C_i=[Q_i',R_i^+,R_{i+1}^-,Q_{i+1}']$ and $R_{i}^+$ and $R_{i}^-$ are the same portions of $R_i$ as defined in \cref{NearTriang} \note{DW}{move in here}    Observe that  $C_i=[ R_i^+ , R_{i+1}^-, I_i ]$, where $R_i^+$ and $R_{i+1}^-$ are bipods, and $I_i$ is the bipod $Q_i' S_i Q_{i+1}'$. Let $G_i$ be the subgraph of $G$ whose vertices and edges are in $C_i$ or its interior.

For $i\in\{1,2,3\}$, if $C_i$ is non-empty, then $G_i$ and $C_i=[ R_i^+ , R_{i+1}^-, I_i ]$ satisfy the conditions of the lemma, and we may apply induction to $G_i$.  (Note that one or two of $R_i^+$, $R_{i+1}^-$ and $I_i$ may be empty, in which case we apply the inductive hypothesis with $k=2$ or $k=1$, respectively.)\  This gives a partition $\PP_i$ of $G_i$ such that $H_i:= G_i / \PP_i$ satisfies the conclusions of the lemma. Let $(B^i_x:x\in V(J_i))$ be a tree-decomposition of $H_i$, in which every bag has size at most 4, and some bag $B^i_{u_i}$ contains the vertices of $H_i$ corresponding to  $R_i^+$, $R_{i+1}^-$ and $I_i$ (if they are non-empty).  We do this for each non-degenerate $C_i$.

We now construct the desired partition $\PP$ of $G$. 
Initialise $\PP:= \{P_1,\ldots,P_k\}$. Then add each non-empty $Q_i'$ to $\PP$. 
Now for each non-degenerate $F_i$, each path in  $\PP_i$ is either an \emph{external} path (that is, fully contained in $F_i$) or is an \emph{internal path} with none of its vertices in $F_i$. 
Add all the internal paths of  $\PP_i$ to $\PP$. 
By construction, $\PP$ partitions $V(G)$ into vertical paths in $T$ and $\PP$ contains $P_1,\ldots,P_k$.

Let $H:=G/\PP$. Next we exhibit the desired tree-decomposition $(B_x:x\in V(J))$ of $H$. 
Let $J$ be the tree obtained from the disjoint union of $J_i$, taken over the $i\in\{1,2,3\}$ such that $F_i$ is non-degenerate, by adding one new node $u$ adjacent to each $u_i$.
(Recall that $u_i$ is the node of $J_i$ for which the bag $B^i_{u_i}$ contains the vertices of $H_i$ corresponding to the paths that form $F_i$.)\ 
Let the bag $B_u$ contain all the vertices of $H$ corresponding to $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$. 
For each non-degenerate $F_i$, and for each node $x\in V(J_i)$, initialise $B_x:= B^i_x$. 
Recall that vertices of $H_i$ correspond to contracted paths in  $\PP_i$.
Each internal path in $\PP_i$ also lies  in $\PP$.
Each external path $P$ in $\PP_i$ is a subpath of $P_j$ for some $j\in\{1,\dots,k \}$ or is one of the paths among $Q'_1, Q'_2, Q'_3$. 
For each such path $P$, for every $x\in V(J)$, in bag $B_x$, 
replace each instance of the vertex of $H_i$ corresponding to $P$ by the vertex of $H$ corresponding to the path among $P_1,\ldots, P_k, Q'_1,\ldots,Q'_3$ 
that contains $P$.
This completes the description of $(B_x : x\in V(J))$. 
By construction, $|B_x|\leq 9$ for every $x\in V(J)$.

First we show that for each vertex $a$ in $H$, the set $X:=\{x\in V(J) : a\in B_x\}$ forms a subtree of $J$.
If $a$ corresponds to a path distinct from $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$ then $X$ is fully contained in $J_i$ for some $i\in\{1,2,3\}$.
Thus, by induction $X$ is non-empty and connected in $J_i$, so it is in $J$.
If $a$ corresponds to $P$ which is one of the paths among $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$ then 
$u\in X$ and whenever $X$ contains a vertex of $J_i$ it is because some external path of $\PP_i$ was replaced by $P$.
In particular, we would have $u_i \in X$ in that case. 
Again by induction each $X\cap J_i$ is connected and since $uu_i\in E(T)$, we conclude that $X$ induces a (connected) subtree of $J$.

Finally we show that, for every edge $ab$ of $H$, there is a bag $B_x$ that contains $a$ and $b$. 
If $a$ and $b$ are both obtained by contracting any of $P_1,\ldots,P_k,Q'_1,Q'_2,Q'_3$, 
then $a$ and $b$ both appear in $B_u$.  
If $a$ and $b$ are both in $H_i$ for some $i\in\{1,2,3\}$, 
then some bag $B^i_x$ contains both $a$ and $b$. 
Finally, when $a$ is obtained by contracting a path $P_a$ in $G_i-V(F_i)$ and $b$ is obtained by contracting a path $P_b$ not in $G_i$, 
then  the cycle $F_i$ separates $P_a$ from $P_b$ so the edge $ab$ is not present in $H$. 
This concludes the proof that $(B_x:x\in V(J))$ is the desired tree-decomposition  of $H$.

We construct $\PP$ as before. Initialise $\PP:=\{P_1,\ldots,P_k,Y\}$. Then, for $i\in \{1,2,3\}$, each tripod in  $\PP_i$ is either fully contained in $F_i$ or it is \emph{internal} with none of its vertices in $F_i$. Add all these internal tripods in  $\PP_i$ to $\PP$. By construction, $\PP$ partitions $V(G)$ into tripods. The graph $H:= G/ \PP$ is planar since $G$ is planar and each tripod in $\PP$ induces a connected subgraph of $G$.

Next we produce the tree-decomposition $(B_x:x\in V(J))$ of $H$ that satisfies the requirements of the lemma.  Let $J$ be the tree obtained from the disjoint union of $J_1$, $J_2$ and $J_3$ by adding one new node $u$ adjacent to $u_1$, $u_2$ and $u_3$. Let $B_u$ be the set of at most four vertices of $H$ corresponding to $Y,P_1,\dots,P_k$. For $i\in\{1,2,3\}$ and for each node $x\in V(J_i)$, initialise $B_x:= B^i_x$.

As in  the proof of \cref{NearTriang},  the resulting structure,  $(B_x:x\in V(J))$, is not yet a tree-decomposition of $H$ since some bags may  contain vertices of $H_i$ that are not necessarily vertices of $H$. Note that unlike in  \cref{NearTriang} this does not only include elements of $\PP_i$  that are contained in $F$. 
In particular, $I_i$ is also not an element of $\PP$ and thus does not correspond to a vertex of $H$.   
We remedy this  as follows. For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $I_i$ by the vertex of $H$ corresponding to $Y$. 
Similarly, by construction, $R_i^+$ is a subgraph of $P_{\alpha_i}$ for some $\alpha_i\in \{1,\dots,k\}$. 
For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $R_i^+$ 
by the vertex of $H$ corresponding to $P_{\alpha_i}$. 
Finally, $R_{i+1}^-$ is a subgraph of $P_{\beta_i}$ for some $\beta_i\in \{1,\dots,k\}$. 
For $x\in V(J)$, in bag $B_x$, replace each instance of the vertex of $H_i$ corresponding to $R_{i+1}^-$  by the vertex of $H$ corresponding to $P_{\beta_i}$.

This completes the description of $(B_x:x\in V(J))$. Clearly, every bag $B_x$ has size at most 4. We now prove that $(B_x:x\in V(J))$  is a tree-decomposition of $H$ 
\end{proof}

\begin{lem}
	An embedded multigraph $G$ is edge-maximal 1-plane with no faces of length 2 if and only if there is a 2-connected plane graph $G_0$ such that: 
	\begin{compactitem}
		\item each face of $G_0$ is bounded by a 3-cycle or a 4-cycle,
		\item no edge of $G_0$ is in the boundary of two facial 3-cycles,
		\item $G=\widehat{G_0}$.
	\end{compactitem}
\end{lem}

\begin{proof}
\end{proof}

\end{document}

%%%%%%%%%%%%%%%%%%%%

I claim that G_2 is contained in H * P * K_{ d * floor(d/2) } where tw(H) <= 3.

With d=4, G_2 includes all edge-maximal 1-planar  multigraphs,
and G_2 is contained in H * P * K_8 where tw(H) <= 3.

Let T be a bfs spanning tree of G_0 rooted on the outerface.

Run the original proof, finding three paths Q_1,Q_2,Q_3
leading to a Sperner triangle t in G_1, which is a subset of a face F of G_0.
The corresponding tripod in our partition is Q_1 \cup Q_2 \cup Q_3 \cup F.
We can apply induction inside the regions bounded by C, Q_1, Q_2, Q_3 and F.

The layering is in G_0. Each edge of G_2 has span at most floor(d/2)
with respect to this layering.

So G_2 is contained in H * P^{floor(d/2)} * K_d, which is contained in
H * P * K_{d floor(d/2) }.

This setup (if it works) also works for map graphs in which every face
is a nation.

I just thought we need to be careful with the vertex 3-colouring.
If v is on the inside of C and the path in T from v to the root
first hits C at some B_i, then what is the colour of v?
This is not a problem there is a noose in G_0 including A_1..A_k, so
the path in T from v to the root must intersect A_1...A_k.

\end{document}

	\begin{figure}[!b]
	\begin{center}
		\begin{tabular}{cc}
			\includegraphics[scale=1.4]{induction-i-3} &
			\includegraphics[scale=1.4]{induction-i-1} \\
			(a) & (b) \\[1em]
			\includegraphics[scale=1.4]{induction-i-2} &
			\includegraphics[scale=1.4]{induction-iii} \\
			(c) & (d) 
		\end{tabular}
	\end{center}
	\caption{The inductive proof of \cref{NearTriang}: (a)~the spanning tree $T$ and the paths $P_1,\ldots,P_4$; 
		(b) the paths $R_1$, $R_2$, $R_3$, and the Sperner triangle $\tau$; 
		(c)~the paths $Q_1'$, $Q_2'$ and $Q_3'$; 
		(d)~the near-triangulations $G_1$, $G_2$, and $G_3$, with the vertical paths of $T$ on $F_1$, $F_2$, and $F_3$.}
	\label{NearTriangProof}
\end{figure}
